embedplyr: Tools for Working With Text Embeddings

### 2025-08-06 paper/paper.Rmd

Dense vector embeddings are the fundamental building block of modern natural language processing [@lauriola2022]. The ability to represent the meaning of texts as a continuous, high dimensional space underlies the recent successes of large language models (LLMs). Embeddings have also revolutionized research methods and inspired new theoretical frameworks in linguistics, psychology, sociology, and neuroscience [see e.g. @duran2019; @feuerriegel2025; @grand2022; @hamilton2018; @kjell2023; @kozlowski2019; @schrimpf2021]. As text embeddings become ubiquitous in the social and behavioral sciences, the need for flexible, easy-to-learn tools increases. Answering this need, **embedplyr** (pronounced "embe-DEE-plier") enables common operations with word and text embeddings within a **tidyverse** [@wickham2019] and/or **quanteda** [@benoit2018] workflow, as demonstrated in @teitelbaum2024. **embedplyr** is designed to facilitate the use of word and text embeddings in common data manipulation and text analysis workflows, without introducing new syntax or unfamiliar data structures to R users.

>>>>

Dense vector embeddings are the fundamental building block of modern natural language processing [@lauriola2022]. The ability to represent the meaning of texts as a continuous, high dimensional space underlies the recent successes of large language models (LLMs). Embeddings have also revolutionized research methods and inspired new theoretical frameworks in linguistics, psychology, sociology, and neuroscience [see e.g. @duran2019; @feuerriegel2025; @grand2022; @hamilton2018; @kjell2023; @kozlowski2019; @schrimpf2021]. As text embeddings become ubiquitous in the social and behavioral sciences, the need for flexible, easy-to-learn tools increases. Answering this need, **embedplyr** (pronounced "embe-DEE-plier") enables common operations with word and text embeddings within a **tidyverse** [@wickham2019] and/or **quanteda** [@benoit2018] workflow, as demonstrated in @teitelbaum2024. Much as **dplyr** is "a grammar of data manipulation, providing a consistent set of verbs that help you solve the most common data manipulation challenges" [@wickham2025], **embedplyr** is a grammar of embeddings manipulation, designed to facilitate the use of word and text embeddings in common analysis workflows without introducing new syntax or unfamiliar data structures to R users.
